---
title: "Coloc_analysis"
author: "Emilia Vartiainen"
date: "6/23/2020"
output: pdf_document
---

```{r libraries, include=FALSE}
library("dplyr")
library("ggplot2")
library("readr")
library("coloc")
library("GenomicRanges")
library("Rsamtools")
library("plyr")
library("data.table")
library("stringr")
```

# Load eQTL catalogues 

``` {r load_data}
tabix_paths = read.delim("https://raw.githubusercontent.com/eQTL-Catalogue/eQTL-Catalogue-resources/master/tabix/tabix_ftp_paths.tsv", 
                         sep = "\t", header = TRUE, stringsAsFactors = FALSE) %>% dplyr::as_tibble()
imported_tabix_paths = read.delim("https://raw.githubusercontent.com/eQTL-Catalogue/eQTL-Catalogue-resources/master/tabix/tabix_ftp_paths_imported.tsv", 
                                  sep = "\t", header = TRUE, stringsAsFactors = FALSE) %>% dplyr::as_tibble()
```

# A general function to quickly import tabix indexed tab-separated files into data_frame

``` {r scanTabixDataFrame}
#' @param tabix_file Path to tabix-indexed text file
#' @param param An instance of GRanges, RangedData, or RangesList
#' provide the sequence names and regions to be parsed. Passed onto Rsamtools::scanTabix()
#' @param ... Additional parameters to be passed on to readr::read_delim()
#' @return List of data_frames, one for each entry in the param GRanges object.

scanTabixDataFrame <- function(tabix_file, param, ...){
  tabix_list = Rsamtools::scanTabix(tabix_file, param = param)
  df_list = lapply(tabix_list, function(x,...){
    if(length(x) > 0){
      if(length(x) == 1){
        #Hack to make sure that it also works for data frames with only one row
        #Adds an empty row and then removes it
        result = paste(paste(x, collapse = "\n"),"\n",sep = "")
        result = readr::read_delim(result, delim = "\t", ...)[1,]
      }else{
        result = paste(x, collapse = "\n")
        result = readr::read_delim(result, delim = "\t", ...)
      }
    } else{
      #Return NULL if the nothing is returned from tabix file
      result = NULL
    }
    return(result)
  }, ...)
  return(df_list)
}
```

# In eQTL Catalogue, variants with multiple rsids are split over multiple rows in the summary statistics files. Thus, we first want to retain only one unique record per variant. To simplify colocalisation analysis, we also want to exclude multi-allelic variants. The following function imports summary statistics from a tabix-index TSV file and performs necessary filtering.

``` {r import_eQTLcatalogue}
import_eQTLCatalogue <- function(ftp_path, region, column_names){
  #Fetch summary statistics with Rsamtools
  summary_stats = scanTabixDataFrame(ftp_path, region, col_names = column_names)[[1]] 
  
  # Should I remove the duplicates?
  # Remove rsid duplicates and multi-allelic variant
  #summary_stats = dplyr::select(summary_stats, -rsid) %>% 
  #  dplyr::distinct() %>% #rsid duplicates
  #  dplyr::mutate(id = paste(chromosome, position, sep = ":")) %>% 
  #  dplyr::group_by(id) %>% 
  #  dplyr::mutate(row_count = n()) %>% dplyr::ungroup() %>% 
  #  dplyr::filter(row_count == 1) #Multialllics
  
  return(summary_stats)
}
```

# Run coloc function 

```{r run_coloc}
#' @param top_loci a data frame with the loci 
#' @param eqtl eQTL_data_name, has a path to the eQTL catalogue
#' @param gwas a path to the gwas data 
#' @param n the number of samples cases + controls
#' @param s for a case control dataset, the proportion of samples in dataset 1 that are cases
#' @param plot logical, if the variants are plotted or not
#' @param endpoint FinnGen endpoint
#' @param eqtl_name eQTL catalogue data set name 
#' @param tissue eQTL catalogue data set tissue 

run_coloc <- function(top_loci, eqtl, gwas, n, s, endpoint, eqtl_name, tissue) {
  
  coloc_df <- data.frame(locus = character(0), finngen_endpoint = character(0), eQTL_dataset = character(0), gene = character(0), 
                         tissue = character(0), nsnps = numeric(0), PP.H0.abf = numeric(0), PP.H1.abf = numeric(0),
                         PP.H2.abf = numeric(0), PP.H3.abf = numeric(0), PP.H4.abf = numeric(0))
  
  for (i in 1:nrow(top_loci)) {
    region_granges = GenomicRanges::GRanges( # define the genomic region 
      seqnames = as.vector(top_loci$chr[i]), 
      ranges = IRanges::IRanges(start = as.vector(top_loci$start[i]), end = as.vector(top_loci$end[i])), 
      strand = "*") 
    
    #region <- paste0(top_loci$chr[i], ":", top_loci$start[i], "-", top_loci$end[i]) # the region as a string
    
    eqtl_region <- import_eQTLCatalogue(eqtl$ftp_path, region_granges, colnames(readr::read_tsv(eqtl$ftp_path, n_max = 1))) # read the eQTL catalogue and filter the variants to the region
    gwas_region <- scanTabix(gwas, index = paste(eqtl, "tbi", sep = "."), param = region_granges) # read the GWAS file and filter the variants to the region 
    gwas_region <- str_replace(gwas_region[[1]], "\t\t", "\tNA\t") # handle empty columns 
    gwas_region <- read.table(text = gwas_region, sep = "\t") # create a data frame 
    gwas_region$variant <- paste(paste0("chr", gwas_region$chrom), gwas_region$pos, gwas_region$ref, gwas_region$alt, sep = "_")
    colnames(gwas_region) <- colnames(read.csv("/Users/eahvarti/Documents/coloc_analysis/Data/finngen_R5_I9_VARICVE_top_loci.tsv", sep = "\t", header = TRUE, nrows = 1))
    colnames(gwas_region)[1] <- "chrom"
    
    for (j in eqtl_region$gene_id) { # Goes through all the genes in eQTL data set and runs coloc for each of them 
      eqtl_subset <- eqtl_region[which(eqtl_region$gene_id == j),] # subset of the eQTL data set based on the gene id 
      gwas_subset <- subset(gwas_region, gwas_region$variant %in% eqtl_subset$variant) # subset of the GWAS data set based on the variant 
      eqtl_subset <- subset(eqtl_subset, eqtl_subset$variant %in% gwas_subset$variant)
      print("Hello 1")
      gwas_subset <- gwas_subset[order(as.vector(gwas_subset$variant)),]
      eqtl_subset <- eqtl_subset[order(as.vector(eqtl_subset$variant)),]
      print("Hello 2")
      colnames(gwas_subset)[which(colnames(gwas_subset) == "variant")] <- "snp"
      colnames(eqtl_subset)[which(colnames(eqtl_subset) == "variant")] <- "snp"
      
      eQTL_dataset = list(pvalues = eqtl_subset$pvalue, 
                      N = (eqtl_subset$an)[1]/2, # Samples size is allele number (AN) divided by 2
                      MAF = eqtl_subset$maf, 
                      type = "quant", 
                      beta = eqtl_subset$beta,
                      snp = eqtl_subset$snp)
      gwas_dataset = list(beta = gwas_subset$beta,
                      varbeta = gwas_subset$sebeta^2, # variance of beta, sebeta = standard error of beta 
                      type = "cc", 
                      snp = gwas_subset$snp,
                      MAF = gwas_subset$maf, 
                      N = n, 
                      s = s) 
      coloc_res = coloc::coloc.abf(dataset1 = eQTL_dataset, dataset2 = gwas_dataset, p1 = 1e-4, p2 = 1e-4, p12 = 1e-5)
      res_formatted = as.data.frame(coloc_res$summary)
      
      results <- c(top_loci$locus_id[i], endpoint, eqtl_name, j, tissue, res_formatted)
      coloc_df <- rbind(coloc_df, results )
    }
  }
  return(coloc_df)
}

```

# Run coloc analysis 

``` {r coloc_analysis}
eQTL_data_name = dplyr::filter(tabix_paths, study == "Lepik_2017", tissue_label == "blood")[1,] # Specify which eQTL dataset is going to be used 

GWAS_path_VARICVE <- c("/Users/eahvarti/Documents/coloc_analysis/Data/finngen_R5_I9_VARICVE.gz")
GWAS_path_ICP <- c("/Users/eahvarti/Documents/coloc_analysis/Data/finngen_R5_O15_ICP.gz")

top_variants_VARICVE <- read.csv("/Users/eahvarti/Documents/coloc_analysis/Data/I9_VARICVE.top.out", sep = "\t", header = TRUE)
top_variants_VARICVE <- top_variants_VARICVE[which(top_variants_VARICVE$chr != 23),]

top_variants_ICP <- read.csv("/Users/eahvarti/Documents/coloc_analysis/Data/O15_ICP.top.out", sep = "\t", header = TRUE)
top_variants_ICP <- top_variants_ICP[which(top_variants_ICP$chr != 23),]

coloc_VARICVE <- run_coloc(top_variants_VARICVE[1,], eQTL_data_name, GWAS_path_VARICVE, n = 207055, s = 17027/207055, endpoint = "Varicose veins", 
                           eqtl_name = "Lepik_2017", tissue = "blood")
coloc_VARICVE <- run_coloc(top_variants_ICP, eQTL_data_name, GWAS_path_ICP, n = 123579, s = 940/123579, endpoint = "Intrahepatic Cholestasis of Pregnancy (ICP)Ê", 
                           eqtl_name = "Lepik_2017", tissue = "blood")
```


# From here on just draft, not important 
--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

# Determine the genomic loci we're using in colc analysis

``` {r region_ranges}
top_variants_VARICVE <- read.csv("/Users/eahvarti/Documents/coloc_analysis/Data/I9_VARICVE.top.out", sep = "\t", header = TRUE)
top_variants_VARICVE <- top_variants_VARICVE[which(top_variants_VARICVE$chr != 23),]

region_granges_VARICVE = GenomicRanges::GRanges(
  seqnames = as.vector(top_variants_VARICVE$chr[1]), 
  ranges = IRanges::IRanges(start = as.vector(top_variants_VARICVE$start[1]), end = as.vector(top_variants_VARICVE$end)[1]), 
  strand = "*") 

top_variants_ICP <- read.csv("/Users/eahvarti/Documents/coloc_analysis/Data/O15_ICP.top.out", sep = "\t", header = TRUE)
top_variants_ICP <- top_variants_ICP[which(top_variants_ICP$chr != 23),]

region_granges_ICP = GenomicRanges::GRanges(
  seqnames = as.vector(top_variants_ICP$chr), 
  ranges = IRanges::IRanges(start = as.vector(top_variants_ICP$start), end = as.vector(top_variants_ICP$end)), 
  strand = "*") 
```


# Determine the eQTL dataset we're using and select the genonic loci, they are in summary_stats variable 

``` {r eQTL_summary_stats}
eQTL_data_name = dplyr::filter(tabix_paths, study == "Lepik_2017", tissue_label == "blood")[1,] # Specify which eQTL dataset is going to be used 

#Extract column names from first file
column_names = colnames(readr::read_tsv(eQTL_data_name$ftp_path, n_max = 1)) 

eQTL_VARICVE = import_eQTLCatalogue(eQTL_data_name$ftp_path, region_granges_VARICVE, column_names) # These are the eQTL summary statistics for "region_granges" genomic loci
eQTL_ICP = import_eQTLCatalogue(eQTL_data_name$ftp_path, region_granges_ICP, column_names)
```

# Read the FinnGen GWAS summary statistics files 

``` {r load_GWAS}
gwas_VARICVE <- fread("/Users/eahvarti/Documents/coloc_analysis/Data/finngen_R5_I9_VARICVE_top_loci.tsv", sep = "\t", header = TRUE)
colnames(gwas_VARICVE)[1:2] <- c("chromosome", "position")
gwas_ICP <- fread("/Users/eahvarti/Documents/coloc_analysis/Data/finngen_R5_O15_ICP_top_loci.tsv", sep = "\t", header = TRUE)
colnames(gwas_ICP)[1] <- "chrm"
```

``` {r run_coloc1}


run_coloc <- function(eqtl_sumstats, gwas_sumstats, n){ # n is the number of samples for GWAS file , get it from a file Sina sent 
  eQTL_dataset = list(pvalues = eqtl_sumstats$pvalue, 
                      N = (eqtl_sumstats$an)[1]/2, # Samples size is allele number (AN) dvided by 2
                      MAF = eqtl_sumstats$maf, 
                      type = "quant", 
                      beta = eqtl_sumstats$beta,
                      snp = eqtl_sumstats$rsid)
  gwas_dataset = list(beta = gwas_sumstats$beta,
                      varbeta = gwas_sumstats$sebeta^2, # variance of beta, have to be the same length with beta  
                      type = "cc", 
                      snp = gwas_sumstats$rsids,
                      MAF = gwas_sumstats$maf, 
                      N = n) 
  coloc_res = coloc::coloc.abf(dataset1 = eQTL_dataset, dataset2 = gwas_dataset, p1 = 1e-4, p2 = 1e-4, p12 = 1e-5)
  res_formatted = dplyr::as_tibble(t(as.data.frame(coloc_res$summary)))
  return(coloc_res)
}

gwas_VARICVE$variant <- paste(paste0("chr", gwas_VARICVE$chromosome), gwas_VARICVE$position, gwas_VARICVE$ref, gwas_VARICVE$alt, sep = "_") 
eqtl_subset <- eQTL_VARICVE[which(eQTL_VARICVE$gene_id == eQTL_VARICVE$gene_id[1]),] # subset of the eQTL data set based on the gene id 
gwas_subset <- subset(gwas_VARICVE, variant %in% eqtl_subset$variant)
eqtl_subset <- subset(eqtl_subset, variant %in% gwas_subset$variant)
      
coloc_VARICVE <- run_coloc(eqtl_subset, gwas_subset, n=17027) 
coloc_ICP <- run_coloc(eQTL_ICP, gwas_ICP, n=940)
```














